Using SMT engine to generate Symbolic Automata\thanks{This work was partially funded by the Associated Team FM4CPS between INRIA and ECNU, Shanghai}

### 2017-04-20 2017-Xudong-SMT/submit-ICTAC2017.tex
one more

\paragraph{Contribution}
In the article we show how:
\begin{itemize}
\item We have refined the open automaton generation algorithm, to make
  it implementable, and we implemented a full working prototype, within the
  VerCors platform. In the process, we had to improve slightly the
  semantics rules, and to add features in the algorithm to deal with
  management of variables and assignements.
  \item We extended the meta-model of action expressions, and the
    associated API, in the VerCors platform, to be able to manage
    parameterisation over the action algebra operators.
  \item We have implemented the interaction between our algorithm and the Z3
    SMT solver, for checking satisfiability of the logical predicates
    generated by the algorithm.
    \item We validated the whole approach on several use-cases, taken
      from two different process languages.
\end{itemize}

@@

\paragraph{Structure.}
In \ref{section:examples} we present two simple examples that will be
running examples to illustrate all parts of the paper. Then in the
background section
\ref{section:pnets} we give a description and a formal definition of
the pNet model, as found in previous publications.
Section \ref{section:op-semantics} recalls the operational semantics
of pNet, including its structural rules, and a sketch of the algorithm
to compute such semantics, but also explains the enhancement we have
done to these definitions.
Section \ref{section:implementation} explains in details our
implementation within the Vercors platform, and shows the full result of
the semantic computation on one of the running examples.
Finally we conclude and discuss perspectives in section
\ref{section:conclusion}.

@@

Several publications \cite{HMZ:PDP15,henrio:Forte2016} already have introduced
many examples of pNets, encoding
operators of various classical process algebras, or more complex
synchronization structures in distributed or parallel languages.
In this paper, we use a running example inspired from Lotos, to illustrate
the main features of pNets and the manner we deal with them in the symbolic
semantics, and in the satifiability checking.

\QIN{
We choose the LOTOS formula P>>(acc(v);Q), composing several operators of LOTOS language.
}

\begin{figure}[h]
  \centerline{\includegraphics[width=8cm]{XFIG/LOTOSFormula.eps}}
  \caption{pNet encodings for the LOTOS formula}  \label{schema:lotos-pnet}
\end{figure}

This example is based on the Enable
operator of the LOTOS language. For the readers not familiar with
Lotos, the pNet encodings here should be sufficient to understand the
Enable operator semantics.

\QIN{
In Fig.~\ref{schema:lotos-pnet}, we show the encoding of the formula using the encoding.
}

\TODO{To be re-written[DONE]}

\QIN{
There are two styles of encodings used in the formula encoding. The first style is ``data oriented'', meaning that the state of the node controller, here is the pLTS C1, is encoded in the value of the pLTS state variable(s). Here we have a single state variable s0, initialized as 0. Its value is ``0'' when the control is on P, and ``1'' when the control is on Q. The second one uses a controller with two states to indicate the position of the control point in the right side of the formula $(acc(v);Q)$. So it is ``state oriented''. That is how we show prefix operator in PN2. 
}
\QIN{
The transitions in the pLTS have the more general syntax: ``<action-expr> [guard] {sequence of assignments}''. Guards and assignments can only use the values of variables of their source state and values from input variables in the action expression; and a transition can only assign variables of its target state. 
}

\QIN{
Each side of the formula is separately act as a hole $P$ and a sub-pNet PN2. Three synchronization vectors are given for synchronizing the behavior of the subnets to make the pNet act as the Enable operator.. The first vector, transmitting an action $a_1$ of P, can only occur if a1 is not a $\delta$, the vector concrete syntax is typically ``$< l, a_1 , \_ > -> a_1$'', in which ``$\_$'' means that the corresponding subnet is not involved in this synchronization. The second one composes the action from the subnets and gives out a synchronized action $\underline{\delta(x)}$, where the underlined action stands for an internal (already synchronized) action, that cannot be further synchronized at upper levels. Synchronized action is a straightforward generalization of the notion of internal actions, that will be convenient for observing internal events during model-checking. The second vector constraints that the transition occurs only when process $P$ performs a $\delta(x)$ action, transferring the control to $Q$, and sending a value $x$ at the same time. This value must be ``accepted''  through an $acc(x)$ action before $Q$ taking the control. The last one acts similar with the the first one, it transmits the action $a_2$ of Q.
}

\QIN{
The sub-pNet PN2 also have two synchronization vectors to make it act as the behavior of Prefix operator.
}

\QIN{
It should be easy to see that this pNet system encodes properly the behavioral semantics of the corresponding CCS expression. Remark that it is built in a systematic structural way, one pNet node encoding each CCS operator. And it is parameterized both at the level of actions variables (occurring within the controllers and the synchronization vectors), and at the level of process parameters (holes).
}

@@

\TODO{I have removed some of the technical stuff, adding referemces to Forte'16.
  We will see if this is enough, and will have to check if this is coherent, and if we can remove more, but for ICTAC, we MUST keep enough formal stuff}

@@

pNets can encode naturally the notion of input actions as found e.g. in value-passing CCS
\cite{Milner89} or of usual point-to-point message passing calculi, but it also allows
for more general mechanisms, like gate negotiation in Lotos, or broadcast
communications.
\TODO{I removed the CCS input action stuff, check later if we need to
  replace it here by something different about input action notation}.

@@

%The preceding definition relies on the auxiliary functions below:

\TODO{we do not use the Sort in the implementation of synch vector
  matching. May be we could remove it from here, but it will not gain
  much...[DONE] --- Well, not really, because you have not chnaged the places where SORT is used, e.g. in definitions 2 and 3 one minimal solution is to add a sentence with a ref to Forte'16 --- but also you have removed the defs of Holes and Leaves, that we really use a lot !}

@@

The \texttt{LOTOS} pNet of Fig. \ref{schema:lotos-pnet} has 2 controllers and 2 holes. We will show its full open Automaton in section \ref{section:full-result}, it has 5 open transitions. We detail one of them here:

@@

Remark the structured states in the conclusion of these rules: both of them start in state $\ostate{00}$, where both $C_1$ and $C_2$ are in their initial states. They end in different structured states, because in $ot_2$ both pLTS conduct transitions, showing that the control has gone from processes $P$ to $(acc(x);Q$, consuming actions $\delta(x)$ and $acc(x)$. Controller $C_1$ has only one state, but its state variable $s_0$ has changed to 1.

@@

When building an Open Automaton as the semantics of a pNet, its
\emph{states}, and the from of the predicates in its
\emph{transitions} have a specific structure:

@@

\begin{figure}[tb]
\begin{mathpar}
  \small
  \inferrule
    {
          \inferrule
              {0 \xrightarrow {\delta}_{C_1} 0 ~~
               [ s_0=0 ]
              }
              {C_1
                \models
                \openrule
                      {0 \xrightarrow{\delta}_{C_1} 0 ~~
                       [ s_0=0 ] ~~
                       \{ s_0 \leftarrow 1\}
                      }
                      {\ostate{0}\xrightarrow{\delta}\ostate{0}}
              }
	\\
      \inferrule
        {
          \inferrule
              {0 \xrightarrow{acc(x)}_{C_2} 1}
              {C_2
                \models
                \openrule
                      {0 \xrightarrow{acc(x)}_{C_2} 1}
                      {\ostate{0}\xrightarrow{acc(x)}\ostate{1}}
              }
        }
        {
          \textrm{acc(x);Q}\models
              \openrule
                  { 0 \xrightarrow {acc(x)}_{C_2} 1}
                  {\ostate{0}\xrightarrow{acc(x)}\ostate{1}}
        }   
    }
    {
     \textrm{PN1}
     \models
        \openrule{
                        0 \xrightarrow{\delta}_{C_1} 0, 0 \xrightarrow{acc(x)}_{C_2} 1 ~~
                       \{\xrightarrow{hb_{12}}_P\} ~~ 
                        [s_0=0 \!\land\! hb_{12}=\delta(x)] ~~ 
                       \{s_0 \xleftarrow{} 1\}           
                      }
     {\ostate{00} \xrightarrow{\tau} \ostate{01}}
      }\vspace{-4ex}
\end{mathpar}
  \caption{Proof of  $ot_{2}$ (with interaction of processes $P$ and action $acc(x)$) for ``P>>(acc(x);Q)''}
  \label{usingrules:ot2}
\end{figure}

\TODO{Change the example...?[DONE]}
\begin{example} \emph{Using the operational rules to compute
    open-transitions}
  In Fig. \ref{usingrules:ot2} we show the deduction tree used to construct and prove the 
  open transition $ot_{2}$ of \texttt{LOTOS}.
%(see example page \pageref{OT:ccs-composed}).
%  \TODO{Eric: change for a deduction tree of CCS expression}
%  The proof tree uses TR1 twice, for the $l$ transition of $C_1$ and for the
%  $b$ transition of $C_2$, then uses an action $b_P$ of hole $P$, and
%  combines the results using the second vector of the PN2 sub-pNet, and
%  the third vector of the top node. This yields a final $\tau$
%  transition. The deduction tree in the figure shows how the
%  predicates are generated in this process.

@@

@@

\subsection{Computing and using open automata}
In this section we present an algorithm to construct the open
automaton representing the behaviour of an open pNet, and we prove that
under reasonable conditions this automaton is finite.

>>>>

\paragraph{Contribution}
In the article we show how:
\begin{itemize}
\item \ERIC{We have refined the open automaton generation algorithm, to make
  it implementable, and we implemented a full working prototype, within the
  VerCors platform. In the process, we had to improve significantly the
  semantics rules, and add features in the algorithm to deal the full
  model, including management of variables and assignements.}
%  \item We extended the meta-model of action expressions, and the
%    associated API, in the VerCors platform, to be able to manage
%    parameterisation over the action algebra operators.
  \item We have implemented the interaction between our algorithm and the Z3
    SMT solver, for checking satisfiability of the logical predicates
    generated by the algorithm.
    \item \ERIC{We validated the whole approach on a small use-case covering all
      important features of the model.}
\end{itemize}

@@

\paragraph{Structure.}
In \ref{section:examples} we present a simple example that will  illustrate all parts of the paper. Then in the
background section
\ref{section:pnets} we give a description and a formal definition of
the pNet model, as found in previous publications.
Section \ref{section:op-semantics} recalls the operational semantics
of pNet, including its structural rules, and a sketch of the algorithm
to compute such semantics, but also explains the enhancement we have
done to these definitions.
Section \ref{section:implementation} explains in details our
implementation within the Vercors platform, and shows the full result of
the semantic computation on the running example.
Finally we conclude and discuss perspectives in section
\ref{section:conclusion}.

@@

Several publications \cite{HMZ:PDP15,henrio:Forte2016} already have introduced
many examples of pNets, encoding
operators of various classical process algebras, or more complex
synchronization structures in distributed or parallel languages.
In this paper, we use a running example inspired from Lotos, to illustrate
the main features of pNets and the way we deal with them in the symbolic
semantics, and in the satifiability checking.

\begin{figure}[h]
  \centerline{\includegraphics[width=8cm]{XFIG/LOTOSFormula.eps}}
  \caption{pNet encodings for the LOTOS expression}  \label{schema:lotos-pnet}
\end{figure}

We choose the expression \texttt{P>>(accept(v);Q)}, that includes the
Enable and Prefix 
operators of the LOTOS language. For the readers not familiar with
Lotos, the pNet encodings here should be sufficient to understand the
Enable operator semantics.
In Fig.~\ref{schema:lotos-pnet}, we show the encoding of the formula using our encoding.

\ERIC{
  This pNet is composed of: a toplevel node PN1 encoding the Enable
  operator, with its ``controller'' C1; a ``hole'' representing the
  process variable P; another node PN2 representing the Prefix
  operator, with its controller C2, and its hole Q. Controllers are
  labelled transition systems (pLTS) with state variables and value-passing
  transitions, defining the dynamics of the operator semantics.}

In order to show all features of the pNet model, we choose to use two
styles of encodings used in the controllers.
The most common style used in process calculi is ``state oriented'',
we use it here for C2: it has two states to
indicate the position of the control point in the right side of the
Prefix expression $(acc(v);Q)$. The second one is
``data oriented'': controler C1 has a single state, with a state
variable s0 encoding the change of control point from left to right sub-net.
The transitions in the pLTS have the more general syntax:
``<action-expr> [guard] {sequence of assignments}''. Guards and
assignments can only use the values of variables of their source state
and values from input variables in the action expression; and a
transition can only assign variables of its target state.

Now look more specifically on the top pNet node PN1. In addition to
its controller it has 2 sub-pNets, the hole P and the sub-pNet PN2.
Three synchronization vectors are given for synchronizing the behavior
of the subnets to make the pNet act as the Enable operator. The first
vector, transmitting an action $a_1$ of P, can only occur if $a_1$ is not
a $\delta$, the vector concrete syntax is typically ``$< l, a_1 , \_ >
-> a_1$'', in which ``$\_$'' means that the corresponding subnet is
not involved in this synchronization, and the ``l'' action is related
to the corresponding transition of the controller C1.
The second vector composes the
action from the subnets and gives out a synchronized action
$\underline{\delta(x)}$, where the underlined action stands for an
internal (already synchronized) action, that cannot be further
synchronized at upper levels. Synchronized action is a straightforward
generalization of the notion of internal actions, that will be
convenient for observing internal events during model-checking. The
second vector constraints that the transition occurs only when process
$P$ performs a $\delta(x)$ action, transferring the control to $Q$,
and sending a value $x$ at the same time. This value must be
``accepted''  through an $acc(x)$ action before $Q$ taking the
control. The last one acts as the first one, but transmitting the
action $a_2$ of Q.  
Similarily the node PN2 has two synchronization vectors to encode
the behavior of the Prefix operator.

It should be easy to see that this pNet system encodes properly the
behavioral semantics of the corresponding Lotos expression. Remark
that it is built in a systematic structural way, one pNet node
encoding each Lotos operator. It is parameterized both at the level of
actions variables (occurring within the controllers and the
synchronization vectors), and at the level of process parameters
(holes).

@@

% \TODO{I have removed some of the technical stuff, adding referemces to Forte'16.
%  We will see if this is enough, and will have to check if this is coherent, and if we can remove more, but for ICTAC, we MUST keep enough formal stuff}

@@

pNets can encode naturally the notion of input actions as found e.g. in value-passing CCS
\cite{Milner89} or of usual point-to-point message passing calculi, but it also allows
for more general mechanisms, like gate negotiation in Lotos, or broadcast
communications.
%\TODO{I removed the CCS input action stuff, check later if we need to
%  replace it here by something different about input action notation}.

@@

By lack of space, we have omitted here the formal (structural)
definition of $Sorts$ (the 
set of actions of a hole or a pNet), $Leaves$ and $Holes$ (all pLTSs
(resp holes) in a pNet hierarchical system). These can be found in \cite{henrio:Forte2016}.
%The preceding definition relies on the auxiliary functions below:

@@

The \texttt{LOTOS} pNet of Fig. \ref{schema:lotos-pnet} has 2
controllers and 2 holes. We will show its full open Automaton in
section \ref{section:full-result}. We detail one of its transition here:

@@

Remark the structured states in the conclusion of this rule, encoding
the combination of state-spaces of the leaves: both of them start in
state $\ostate{00}$, where both $C_1$ and $C_2$ are in their initial
states. They end in different structured states, because in $ot_2$
both pLTS conduct transitions, showing that the control has gone from
processes $P$ to $(acc(x);Q$, consuming actions $\delta(x)$ and
$acc(x)$. Controller $C_1$ has only one state, but its state variable
$s_0$ has changed to 1.

@@

When building an Open Automaton as the semantics of a pNet, its
\emph{states}, and the shape of the predicates in its
\emph{transitions} have a specific structure:

@@

\begin{example} \emph{Using the operational rules to compute
    open-transitions}
  In Fig. \ref{usingrules:ot2} we show the deduction tree used to construct and prove the 
  open transition $ot_{2}$ of \texttt{LOTOS}.
%(see example page \pageref{OT:ccs-composed}).
%  \TODO{Eric: change for a deduction tree of CCS expression}
%  The proof tree uses TR1 twice, for the $l$ transition of $C_1$ and for the
%  $b$ transition of $C_2$, then uses an action $b_P$ of hole $P$, and
%  combines the results using the second vector of the PN2 sub-pNet, and
%  the third vector of the top node. This yields a final $\tau$
%  transition. The deduction tree in the figure shows how the
%  predicates are generated in this process.

@@

\begin{figure}[h]
\begin{mathpar}
  \small
  \inferrule
    {
          \inferrule
              {0 \xrightarrow {\delta}_{C_1} 0 ~~
               [ s_0=0 ]
              }
              {C_1
                \models
                \openrule
                      {0 \xrightarrow{\delta}_{C_1} 0 ~~
                       [ s_0=0 ] ~~
                       \{ s_0 \leftarrow 1\}
                      }
                      {\ostate{0}\xrightarrow{\delta}\ostate{0}}
              }
	\\
      \inferrule
        {
          \inferrule
              {0 \xrightarrow{acc(x)}_{C_2} 1}
              {C_2
                \models
                \openrule
                      {0 \xrightarrow{acc(x)}_{C_2} 1}
                      {\ostate{0}\xrightarrow{acc(x)}\ostate{1}}
              }
        }
        {
          \textrm{acc(x);Q}\models
              \openrule
                  { 0 \xrightarrow {acc(x)}_{C_2} 1}
                  {\ostate{0}\xrightarrow{acc(x)}\ostate{1}}
        }   
    }
    {
     \textrm{PN1}
     \models
        \openrule{
                        0 \xrightarrow{\delta}_{C_1} 0, 0 \xrightarrow{acc(x)}_{C_2} 1 ~~
                       \{\xrightarrow{hb_{12}}_P\} ~~ 
                        [s_0=0 \!\land\! hb_{12}=\delta(x)] ~~ 
                       \{s_0 \xleftarrow{} 1\}           
                      }
     {\ostate{00} \xrightarrow{\tau} \ostate{01}}
      }\vspace{-4ex}
\end{mathpar}
  \caption{Proof of  $ot_{2}$ (with interaction of processes $P$ and action $acc(x)$) for ``P>>(acc(x);Q)''}
  \label{usingrules:ot2}
\end{figure}

@@

\subsection{Computing and using open automata}
In this section we present an algorithm to construct the open
automaton representing the behavior of an open pNet, and we prove that
under reasonable conditions this automaton is finite.
